---
title: "Time Series Forecasting In R"
author: "Conor Tompkins"
format:
  revealjs:
    embed-resources: true
editor: visual
---

## What is time series forecasting?

Time series forecasting is the practice of making predictions about the future value of some quantitative variable.

Forecast accuracy is typically the focus, not inference.

## What is a time series?

Data collected about a quantitative variable sequentially over time.

-   Customer demand for a company's product

-   Price of electricity

-   Stock price

-   Pothole complaints

## What can be forecast?

Factors include:

1.  how well we understand the factors that contribute to it;

2.  how much data is available;

3.  how similar the future is to the past;

4.  whether the forecasts can affect the thing we are trying to forecast.

## Why forecast?

Improve decision making about the future.

Examples

1.  a

2.  b

3.  c

## Forecast horizons

-   Short term

    -   How many clerks will a grocery store need tomorrow?

-   Medium term

    -   How many flowers will the flower shop need to buy to fulfill demand for Valentine's Day?

-   Long term

    -   How many retail stores will a company need to open to meet future demand given changing population trends across geography?

## How?

#### Typical regression model

Pothole complaints as a function of explanatory variables

```{r}
#| eval: false
#| echo: true
pothole_complaints ~ year + month + weather + public_works_budget + error
```

#### Time series model

Future value of pothole complaints as a function of the previous values of pothole complaints (plus explanatory variables)

```{r}
#| eval: false
#| echo: true
pothole_complaints(t+1) ~ pothole_complaints(t) + pothole_complaints(t-1) + pothole_complaints(t-2) + ... + error
```

## Forecasting Process

1.  Exploratory data analysis

2.  Model selection

3.  Forecast

4.  Evaluate forecast accuracy

## Pothole walkthrough

Read in the data

## Pothole walkthrough

Data structure

## Pothole walkthrough

Split 80/20 train/test

## Exploratory data analysis

Basic time series plot

## Exploratory data analysis

Decomposition

## Model selection

Explain various basic models

## Forecast

## Evaluate forecast accuracy

## Appendix

Cross-validation

## Appendix

Exogenous variables
