---
title: "Time Series Forecasting In R"
author: "Conor Tompkins"
format:
  revealjs:
    embed-resources: true
editor: visual
---

## What is time series forecasting?

Time series forecasting is the practice of making predictions about the future value of some quantitative variable.

Forecast accuracy is typically the focus, not inference.

## What is a time series?

Data collected about a quantitative variable sequentially over time.

-   Customer demand for a company's product

-   Price of electricity

-   Stock price

-   Pothole complaints

## What can be forecast?

Variables that affect forecastability:

1.  Do we understand the underlying process that creates the time series (data generating process)?

2.  Is historical data is available?

3.  Will the future be similar to the past?

4.  Will the forecast affect the thing we are trying to forecast?

## What can be forecast?

"How much will it rain tomorrow"

vs.

"How much rain will fall on you tomorrow"

## Why forecast?

Improve decision making about the future.

## Forecast horizons

-   Short term

    -   How many clerks will a grocery store need tomorrow?

-   Medium term

    -   How many flowers will the flower shop need to buy to fulfill demand for Valentine's Day?

-   Long term

    -   How many retail stores will a company need to meet future demand given changing population trends across geography?

## How?

#### Typical regression model

Pothole complaints as a function of explanatory variables

```{r}
#| eval: false
#| echo: true
pothole_complaints ~ year + month + weather + public_works_budget + error
```

#### Time series model

Future value of pothole complaints as a function of the previous values of pothole complaints (plus explanatory variables)

```{r}
#| eval: false
#| echo: true
pothole_complaints(t+1) ~ pothole_complaints(t) + pothole_complaints(t-1) + pothole_complaints(t-2) + ... + error
```

## Forecasting Process

1.  Exploratory data analysis

2.  Model selection

3.  Forecast

4.  Evaluate forecast accuracy

## Pothole walkthrough

Read in the data

## Pothole walkthrough

Data structure

## Pothole walkthrough

Split 80/20 train/test

## Exploratory data analysis

Basic time series plot

## Exploratory data analysis

Decomposition

## Modelling

Train/test split

## Modelling

Explain various basic models

## Modelling

Forecast

## Modelling

Evaluate forecast accuracy

## Appendix

Cross-validation

## Appendix

Exogenous variables
