---
title: "Time Series Forecasting In R"
author: "Conor Tompkins"
format:
  revealjs:
    embed-resources: true
editor: visual
---

## What is time series forecasting?

Time series forecasting is the practice of making predictions about the future value of some quantitative variable.

Forecast accuracy is typically the focus, not inference.

## What is a time series?

Data collected about a quantitative variable sequentially over time.

-   Customer demand for a company's product

-   Price of electricity

-   Stock price

-   Pothole complaints

Miniature graph of a time series goes here

## What can be forecast?

Forecastability is determined by:

1.  Do we understand the underlying process that creates the time series (data generating process)?

2.  Is historical data is available?

3.  Will the future be similar to the past?

4.  Will the forecast affect the thing we are trying to forecast?

    -   Feedback loop

## What can be forecast?

> "How much will it rain tomorrow"

vs.

> "How much rain will fall on you tomorrow"

## Why forecast?

-   Improve decision making about the future.

-   Typically related to resource allocation.

## Forecast horizons

-   Short term

    -   What will be the peak demand for electricity be in the next hour?

-   Medium term

    -   How many flowers will the flower shop need to buy to fulfill demand for Valentine's Day?

-   Long term

    -   What will future demand for a company's products be given changing population trends across geography?

## How?

#### Typical regression model

Pothole complaints as a function of explanatory variables

```{r}
#| eval: false
#| echo: true
pothole_complaints ~ year + month + weather + public_works_budget + error
```

#### Time series model

Future value of pothole complaints as a function of the previous values of pothole complaints (plus explanatory variables)

```{r}
#| eval: false
#| echo: true
pothole_complaints(t+1) ~ pothole_complaints(t) + pothole_complaints(t-1) + pothole_complaints(t-2) + ... + error
```

## Forecasting Process

1.  Exploratory data analysis

2.  Model selection

3.  Forecast

4.  Evaluate forecast accuracy

## Pothole walkthrough

Read in the data

```{r}
library(fpp3)
library(readr)
library(janitor)
library(future)

custom_theme <- theme_bw() + 
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 18))

theme_set(custom_theme)

plan(multisession)

options(scipen = 999, digits = 4)

set.seed(1234)

#read in pothole data
#https://data.wprdc.org/datastore/dump/29462525-62a6-45bf-9b5e-ad2e1c06348d
report_data <- read_csv("inputs/wprdc_311_2024_10_20.csv") |> 
  clean_names() |>
  mutate(create_date = yearmonth(create_date_et)) |> 
  rename(request_type = request_type_name)
```

## Pothole walkthrough {.smaller}

::::: columns
::: {.column width="50%"}
Time series table `({tsibble})` data structure:

-   Key: `request_type`

-   Index: `create_date`

-   Measured variable: `report_count`
:::

::: {.column width="50%"}
```{r}
#create basic tsibble
pothole_df <- report_data |> 
  filter(request_type == "Potholes") |> 
  summarize(report_count = n(),
            .by = c(create_date, request_type)) |> 
  ungroup() |>
  filter(year(create_date) >= 2016) |> 
  as_tsibble(key = request_type, index = create_date)

pothole_df
```
:::
:::::

## Exploratory data analysis

```{r}
autoplot(pothole_df)
```

## Exploratory data analysis

Seasonal plot showing each month across years

```{r}
gg_season(pothole_df)
```

## Exploratory data analysis

Seasonal subseries plot

```{r}
gg_subseries(pothole_df) +
  facet_wrap(vars(month(create_date, label = TRUE)), ncol = 3)
```

## Exploratory data analysis

Time series decomposition

```{r}
dcmp <- pothole_df |>
  model(stl = STL(report_count, robust = TRUE))

dcmp_components <- components(dcmp)

autoplot(dcmp_components)
```

## Exploratory data analysis

Outliers

```{r}
outliers <- dcmp_components |>
  filter(remainder < quantile(remainder, 0.25) - 3*IQR(remainder) |
           remainder > quantile(remainder, 0.75) + 3*IQR(remainder))

pothole_df |>
  ggplot(aes(create_date, report_count)) +
  geom_line() +
  geom_point(data = outliers, color = "red")
```

## Modelling

-   Train/test split

-   Use last 20% of observations as test data

```{r}
#| echo: true
data_test <- pothole_df |> 
  slice_tail(prop = .2)

data_train <- pothole_df |> 
  anti_join(data_test, by = "create_date")
```

## Modelling {.smaller}

::::: columns
::: column
-   NAIVE : play the last value forward over the forecast horizon

-   SNAIVE: play the last seasonal value forward

-   MEAN : take the average of the entire series and play it forward

    -   Use `window()` to apply a window for moving averages
:::

::: column
-   `TSLM` : fit a linear model

    -   Use `seasonal() to add`dummy variables for seasonal effects

-   ETS : weighted average based on recency of observation

-   ARIMA :
:::
:::::

## Modelling

Fit models on training data

```{r}
model_df <- data_train |> 
  model(naive = NAIVE(log(report_count + 1)),
        naive_seasonal = SNAIVE(log(report_count + 1)),
        mean = MEAN(log(report_count + 1)),
        mean_moving_6 = MEAN(log(report_count + 1), window = 6),
        mean_moving_12 = MEAN(log(report_count + 1), window = 12),
        lm = TSLM(log(report_count + 1) ~ trend()),
        lm_seasonal = TSLM(log(report_count + 1) ~ trend() + season()),
        arima = ARIMA(log(report_count + 1)),
        ets = ETS(log(report_count + 1)))

model_df
```

## Modelling

Make forecast from fit models

```{r}
pothole_fc <- model_df |> 
  forecast(data_test)

pothole_fc
```

## Modelling

-   Metrics:

    -   Root Mean Squared Error (RMSE)

    -   Continuous Ranked Probability Score (CRPS)

        -   Measures how well the forecast distribution fits the test data

        -   "skill" measures CRPS compared to a naive benchmark model

    -   Others available in `{fabletools}`

## Modelling

Evaluate forecast accuracy based on full time series (train + test)

```{r}
fc_acc <- pothole_fc |> 
  accuracy(pothole_df,
           measures = list(point_accuracy_measures, distribution_accuracy_measures, skill_cprs = skill_score(CRPS))) |> 
  rename(rmse = RMSE) |> 
  select(request_type, .model, .type, skill_cprs, rmse) |> 
  arrange(desc(skill_cprs))

fc_acc
```

## Modelling

Plot forecast vs test data

```{r}
model_acc <- fc_acc |> 
  pull(.model)

pothole_fc <- pothole_fc |> 
  mutate(.model = factor(.model, levels = model_acc))

pothole_fc |> 
  mutate(.model = factor(.model, levels = model_acc)) |> 
  filter(.model %in% model_acc[1:3]) |> 
  autoplot(data = pothole_df |> filter(year(create_date) >= 2021)) +
  facet_wrap(vars(.model), scales = "free_y", ncol = 1) +
  guides(fill_ramp = "none",
         fill = "none",
         color = "none") +
  labs(title = "Top 3 forecasts by model",
       subtitle = "Sorted descending by accuracy")
```

## Modelling

Inspect top model

```{r}
model_df |> 
  select(lm_seasonal) |> 
  report()
```

## Modelling

Refit top model on entire time series and make a true 12 month forecast

```{r}
final_model <- pothole_df |> 
  model(lm_seasonal = TSLM(log(report_count + 1) ~ trend() + season()))

final_model |> 
  forecast(h = 12) |> 
  autoplot(pothole_df |> filter(year(create_date) >= 2021)) +
  labs(title = "Final 12 month forecast of pothole complaints",
       x = "Report create date",
       y = "Report count")
```

## Appendix

Cross-validation

## Appendix

Exogenous variables
